Pythran: Enabling Static Optimization of Scientific Python Programs

### 2014-01-28 doc/papers/iop2014/paper.tex
X. Corbillon proofread
Update IOP2014 paper with merge from Adrien's diff

    Pythran is an open source static compiler that turns modules written
    in a subset of Python into native ones. Based on the fact that scientific
    modules do not rely much on the dynamic features of the language, it trades
    them in favor of powerful, eventually inter procedural, optimizations.
    These include detection of pure functions, temporary allocation removal,
    constant folding, Numpy \texttt{ufunc} fusion and parallelization, explicit
    thread-level parallelism through OpenMP annotations, false variable
    polymorphism pruning, and automatic vector instruction generation such as
    AVX or SSE.

    In addition to these compilation steps, Pythran provides a C++ runtime library that
    leverages the C++ STL to provide generic containers, and the Numeric
    Template Toolbox (NT2) for Numpy support. It takes advantage of modern C++11
    features such as variadic templates, type inference, move semantics and
    perfect forwarding, as well as classical ones such as expression templates.
    % FIXME Expression template is not a "feature" of the language IMO

    The input code remains compatible with the Python interpreter, and output
    code is generally as efficient as the annotated Cython equivalent, if not
    more, without the backward compatibility loss of Cython. Numpy expressions
    run faster than when compiled with numexpr or numba, without any change to
    the original code.

@@

The Python language is growing in popularity as a language for scientific
computing, mainly thanks to a concise syntax, a high level standard library and
several scientific packages.

However, the overhead of running a scientific application written in Python
compared to the same algorithm written in a statically compiled language such
as C is high, due to numerous dynamic lookup and interpretation cost inherent
in high level languages. Additionally, the Python compiler performs no
optimization on the byte code, while scientific applications are first-class
candidates for many of them.

Following the saying that scientific applications spend 90\% of their time in
10\% of the code, it is natural to focus on computation-intensive piece of code.
The aim may not be to optimize the full Python application, but rather a
small subset of the application.

Several tools have been proposed by an active community to fill the performance
gap encountered when running these computation-intensive piece of code, either
through static compilation or Just In Time (JIT) compilation.

An approach used by Cython\cite{cython2010} is to suppress the interpretation
overhead by translating Python Programs to C programs calling the Python C
API\cite{pythoncapi}. More recently, Nuitka\cite{nuitka}  has taken the same
approach using C++ has a back-end. Going a step further Cython also provides an
hybrid C/Python language that can efficiently be translated to C code, relying
on the Python C API for some parts and on plain C for others.
ShedSkin\cite{shedskin2006} translates implicitly strongly typed Python program
into C++, without any call to the Python C API, but fails to compile dynamic
codes or codes for which static type inference is impossible.

The alternate approach consists in writing a JIT compiler, embedded into the
interpreter, to dynamically turn the computation intensive parts into native
code. The \texttt{numexpr} module\cite{numexpr} does so for Numpy expressions
by JIT-compiling them from a string representation to native code.
Numba\cite{numba} and Parakeet\cite{parakeet2012} extend this approach to
Numpy-centric applications while PyPy\cite{pypy2009} applies it to the whole
language, including the Numpy module through the in-progress \texttt{numpypy}
branch.

To the notable exception of PyPy, these compilers do not apply any of the
static optimization techniques that have been known for decades and
% FIXME And Numba? It generates LLVM, which uses static optimization techniques
successfully applied to statically compiled language such as C or C++.
Translators to statically compiled languages do take advantage of them
indirectly, but the quality of generated code may prevent advanced
optimizations, such as vectorization, while they are available at higher level,
i.e. at the Python level. Taking into account the specificities of the Python
language can unlock many new transformations. For instance, PyPy automates the
conversion of the \texttt{range} builtin into \texttt{xrange} through the use
of a dedicated structure called \texttt{range-list}.

@@

Pythran is a compiler for a subset of the Python language. In this paper, the
name \textbf{Pythran} will be used indifferently to refer to the language or
the associated compiler. The input of the Pythran compiler is a Python module
---not a Python program--- meant to be turned into a native module. Typically,
computation-intensive parts of the program are moved to a module fed to
Pythran. The module is translated to polymorphic C++ code through the
generation of templated classes in place of Python functions, then these
classes are either used from regular C++ code or instantiated for given types
to generate a native Python module.

It is a strong requirement for Pythran to maintain backward compatibility with
Python, that is unlike Cython, all code compilable by Pythran are still
executable by a standard-conforming Python interpreter. For that reason, in
addition to language restrictions detailed in the following, Pythran
understands special comments such as:

@@

as optional module signature. One does not need to list all the module
functions in an \texttt{export} directive, only the functions meant to be used
outside of the module. Polymorphic functions can be listed several times with
different types. These type annotations are only used for native code generation.

@@

    \item turn Python code into Python Abstract Syntax Tree (AST) thanks to the \texttt{ast}
   module from the standard library;

@@

Pythran IR is similar to Python AST, as defined in the \texttt{ast} module, except
that several nodes are forbidden (most notably Pythran does not support
user-defined classes, or the \texttt{exec} instruction), and some nodes are converted
to others to form a simpler AST easier to deal with for further analyses and
optimizations. The transformations applied by Pythran on Python AST are the
following:

@@

    \item the closure of nested functions is statically computed to turn nested
        functions into global functions taking the closure elements as
        parameters, and the initial definition is replaced by a \emph{partial
        function} creation, using the \texttt{partial} type from the standard \texttt{functools} module;

@@

    \item all imports are fully expanded to make function and global variables access paths explicit,

@@

    \item \texttt{try \dots finally} constructs are turned into nested \texttt{try \dots except} blocks;

    \item identifiers whose name may clash with C++ keywords are renamed.

@@

The back-end works in three steps:

@@

The first step requires to map polymorphic variables and polymorphic functions
from the Python world to C++. Pythran only supports polymorphic variables for
functions, i.e. a variable can hold several function objects during its life
time, but it cannot be assigned to a string if it has already been assigned to
an integer. As shown later, it is possible to detect several false variable
polymorphism cases using use-def chains. Function polymorphism is achieved
through template parameters: a template function can be applied to several
types as long as an implicit structural typing is respected, which is very
similar to Python's duck typing, except that it is checked at compile time, as
illustrated by the following implementation of a generic dot product in Python:

@@

Although far more verbose than the Python version, the C++ version also uses a
form of structural typing: the only assumption these two versions make are that
\texttt{L0} and \texttt{L1} are iterable, their content can be multiplied and the result of
the multiplication is accumulatable.

Finally, Pythran computes the set of functions and types used for the processed
module, and generate the minimal set of headers that need to be included to
provide the definitions needed to compile the generated C++ code. This step is
critical to prevent typical slowdown from C++ compilation.

The second step only consists in the instantiation of the top-level functions of
the module, using user-provided signatures. Template instantiation then triggers
the different correctly typed instantiations for all functions written in the
module. Note that the user only needs to provide the type of the functions
exported outside the module. The possible types of all internal functions are
then inferred from the call sites. Note that this complex step leverage the 
C++11 compiler capabilities: Pythran generates only one line using user-provided
signature to trigger the template instanciation by the C++ compiler.

The last step involves a template library, called \texttt{pythonic} that 
contains a polymorphic implementation of many functions from the Python standard
library in the form of C++ template functions. Several optimizations, most
notably using expression template, are delegated to this library. Pythran relies
on the C++11\cite{isocxx11} language, as it makes heavy use of recent features 
such as move semantics, type inference through \texttt{decltype(\dots)} and 
variadic templates.
As a consequence it requires a C++11 compatible compiler for the native code
generation. Boost.Python\cite{boostpython2007} is involved for the automatic
Python-to-C++ boilerplate. Generated C++ code is compatible starting with g++ 
4.7.2 and clang++ 3.2.

It is important to note that all Pythran analyses and optimizations are
type-agnostic, i.e. they do not assume any type for the variables manipulated
by the program. The generated C++ template is still type agnostic, the specialization occurs only by generating the template instanciation.
Said otherwise, the Pythran compiler analyzes polymorphic functions and turns
Python modules into C++ meta-programs.

@@

A code analysis is a function that takes a part of the IR (or the whole
module's IR) as an input and returns aggregated high-level information. For
instance, a simple Pythran analysis called \texttt{Identifiers} gathers the set
of all identifiers used throughout the program. This information is later used
when the creation of new identifiers is required so that no conflict occurs
with existing ones.

@@

As we do not specialise the analysis for different types and the truth value of
\texttt{a} is unknown at compilation time, the alias analysis yields the approximated
result that \texttt{c} may point to a fresh location, \texttt{a} or \texttt{b}.

Without this kind of information, even a simple instruction like
\texttt{sum(a)} would yield very few informations as there is no guarantee that
the \texttt{sum} identifiers points to the \texttt{sum} built-in, as someone
could have previously stated that \texttt{sum = len}.

When turning Python AST to Pythran IR, nested functions are turned into global
functions taking their expanded closure as parameter. This closure is computed using the
information provided by the \texttt{Globals} analysis that statically computes the
state of the dictionary of globals, and \texttt{ImportedIds} that computes the set of
identifiers used by an instruction but not declared in this instruction. For
instance in the following snippet:

@@

A rather high-level analysis is the \texttt{PureFunctions} analysis, that computes the
set of functions declared in the module that are pure, i.e. whose return value
only depends from the value of their argument. This analysis depends on two
other analyses, namely \texttt{GlobalEffects} that computes for each function whether
this function modifies the global state (including I/O, random generators, etc.)
and \texttt{ArgumentEffects} that computes for each argument of each function whether
this argument may be updated in the function body. These three analyses work
inter-procedurally, as illustrated by the following example:

@@

The \texttt{fibo} function is pure as it has no global effects or argument effects and
only calls itself. As a consequence the \texttt{bar} function is also pure as the
\texttt{map} intrinsic is pure when its first argument is pure. However the \texttt{foo}
function is not pure as it calls the \texttt{sample} function from the \texttt{random}
module, which has a global effect (on the underlying random number generator
internal state).

Several analyses depend on the \texttt{PureFunctions} analysis.
\texttt{ParallelMaps} uses aliasing information to check if an identifier
points to the \texttt{map} intrinsic, and checks if the first argument is a
pure function using \texttt{PureFunctions}. In that case the \texttt{map} is
added to the set of parallel maps, because it can be executed in any order.
This is the case for the first \texttt{map} in the following snippet, but not
for the second because the \texttt{print b} involves an \textbf{I/O}.

@@

\texttt{ConstantExpressions} uses function purity to decide whether a given expression
is constant, i.e. its value only depends on literals. For instance the
expression \texttt{fibo(12)} is a constant expression because \texttt{fibo} is pure and its
argument is a literal.

\texttt{UseDefChains} is a classical analysis from the static compilation world. For
each variable defined in a function, it computes the chain of \textbf{use} and \textbf{def}.
The result can be used to drive various code transformations, for instance to
remove dead code, as a \textbf{def} followed by a \textbf{def} or nothing is useless. It is
used in Pythran to avoid false polymorphism. An intuitive way to represent
use-def chains is illustrated on next code snippet:

@@

The fact that all paths finish by a \textbf{def} indicates that the last
assignment can be removed (but not necessarily its right hand part that could
have a side-effect). The fact that the last \textbf{use} is always preceded by
a \textbf{def} that only point to it indicates that these \textbf{def}s can
safely be renamed from \texttt{a} to \texttt{a}' (thus performing scalar
renaming).
% FIXME The scalar renaming part is unclear

All the above analyses are used by the Pythran developer to build code
transformations that improve the execution time of the generated code.

@@

However, there are some informations available at the Python level that cannot
be recovered at the C++ level. For instance, Pythran uses functor with an
internal state and a goto dispatch table to represent generators. Although
effective, this approach is not very efficient, especially for trivial cases.
Such trivial cases appear when a generator expression is converted, in the
front-end, to a looping generator. To avoid this extra cost, Pythran turns
generator expressions into call to \texttt{imap} and \texttt{ifilter} from the
\texttt{itertools} module whenever possible, removing the unnecessary goto
dispatching table. This kind of transformation cannot be made by the C++
compiler. For instance, the one-liner \texttt{len(set(vec[i]+i for i in cols))}
extracted from the \texttt{nqueens} benchmarks from the Unladen Swallow project
is rewritten as \texttt{len(set(itertools.imap(lambda i: vec[i]+i,cols)))}.
This new form is less efficient in pure Python (it implies one extra function
call per iteration), but can be compiled into C++ more efficiently than a
generic generator.

@@

The information concerning constant expressions is used to perform a classical
transformation called \texttt{ConstantFolding}, which consists in the compile-time
evaluation of constant expressions. The validity is guaranteed by the
\texttt{ConstantExpressions} analysis, and the evaluation relies on Python ability to
compile an AST into byte code and run it, benefiting from the fact that Pythran
IR is a subset of Python AST. A typical illustration is the initialization of a
cache at compile-time:

@@

Sometimes, coders use the same variable in a function to represent value with
different types, which leads to false polymorphism, as in:

@@

Using the proper library, the C++ language provides an abstraction level close
to what Python proposes. Pythran provides a wrapper library, \texttt{pythonic},
that leverage on the C++ Standard Template Library (STL), the GNU Multiple
Precision Arithmetic Library (GMP) and the Numerical Template Toolbox
(NT2)\footnote{\url{https://github.com/MetaScale/nt2}} to emulate Python
standard library. The STL is used to provide a typed version of the standard
containers (\texttt{list}, \texttt{set}, \texttt{dict} and \texttt{str}), as
well as reference-based memory management through \texttt{shared\_ptr}. Generic
algorithms such as \texttt{accumulate} are used when possible. GMP is the
natural pick to represent Python's \texttt{long} in C++. NT2 provides a generic
vector library called \texttt{boost.simd}\cite{esterie2012boost} that enables
the vector instruction units of modern processors in a generic way. It is used
to efficiently compile Numpy expressions.

@@

    \item Expression templates\cite{expression_templates} are used to avoid
        multiple iterations and the creation of intermediate arrays. Because
        they aggregates all \texttt{ufunc} into a single expression at compile
        time, they also increase the computation intensity of the loop body,
        which increases the impact of the two following optimizations.

@@

When benchmarked with the \texttt{timeit} module on an hyper-threaded quad-core i7, the
pure Python execution yields:

@@

Many scientific applications can benefit from the parallel execution of their
kernels. As modern computers generally feature several processors and several
cores per processor, it is critical for the scientific application developer to
be able to take advantage of them.

@@

The OpenMP standard\cite{openmp3.1} is a widely used solution for Fortran, C
and C++ to describe loop-based and task-based parallelism. It consists of a few
directives attached to the code, that describe parallel loops and parallel code
sections in a shared memory model.

@@

@@

All benchmarks presented in this section are run on an hyper-threaded quad-core
i7, using examples shipped along with Pythran sources, available at
https://github.com/serge-sans-paille/pythran in the \texttt{pythran/test/cases}
directory. The Pythran version used is the \texttt{HEAD} of the \texttt{scipy2013} branch,
ShedSkin 0.9.2, PyPy 2.0 compiled with the \texttt{-jit} flag, CPython 2.7.3, Cython
0.19.1 and Numexpr 2.0.1. All timings are made using the \texttt{timeit} module,
taking the best of all runs. All C++ codes are compiled with g++ 4.7.3, using
the tool default compiler option, generally \texttt{-O2} plus a few optimizing flags
depending on the target.

Cython is not considered in most benchmarks, because to get an efficient
binary, one needs to rewrite the original code, while all the considered tools
are running the very same Python code that remains compatible with CPython. The
experiment was only done to have a comparison with Numexpr.

@@

It comes at no surprise that all tools get more than decent results on this
benchmark. PyPy generates a code almost as efficient as ShedSkin. Altough both
generate C++, Pythran outperforms ShedSkin thanks to a higher level generated
code. For instance all arrays are represented in ShedSkin by pointers to arrays
that likely disturbs the g++ optimizer, while Pythran uses a vector class wrapping
shared pointers.

@@

It seems that compilers have difficulties to take advantage of high level
constructs such as generator expressions, as the overall speedup is not
breathtaking. Pythran benefits from the conversion to \texttt{itertools.imap} here,
while ShedSkin and PyPy rely on more costly constructs. A deeper look at the
Pythran profiling trace shows that more than half of the execution time is
spent allocating and deallocating a \texttt{set} used in the internal loop. There is a
memory allocation invariant that could be taken advantage of there, but none of
the compiler does.

@@

The speed ups are not amazing for a numerical application. there are two
reasons for this poor speedups. First, the \texttt{hyantes} benchmark makes heavy
usage of trigonometric functions, and there is not much gain there. Second, and
most important, the benchmark produces a big 2D array stored as a list of list,
so the application suffers from the heavy overhead of converting them from C++
to Python. Running the same benchmark using Numpy arrays as core containers
confirms this assumption, as illustrated by Table~\ref{tbl:np-hyantes}. This
table also demonstrates the benefits of manual parallelization using OpenMP.

@@

\begin{table}
    \centering
    \begin{tabular}{|l|c|c|c|c|}
    \hline
     Tool    &  CPython    &  Cython  &  Numexpr    & Pythran   \\
    \hline
     Timing  &  192.2ms    &  36.0ms  &    41.2ms   &  17.1ms   \\
    \hline
     Speedup &  x1         &  x5.33   &  x4.67      &  x11.23   \\
    \hline
\end{tabular}
\caption{Benchmarking result on the arc distance kernel.}
\label{tbl:arc-distance}
\end{table}

@@

Although Pythran focuses on a subset of Python and its standard library, many
optimizations opportunities are still possible. Using a Domain Specific
Language(DSL) approach, one could use rewriting rules to optimize several
Python idioms. For instance, \texttt{len(set(x))} could lead to an optimized
\texttt{count\_uniq} that would iterate only once on the input sequence.

>>>>

    Pythran is an open source static compiler that turns modules written
    in a subset of Python language into native ones. Assuming that scientific
    modules do not rely much on the dynamic features of the language, it trades
    them for powerful, eventually inter procedural, optimizations.
    These optimizations include detection of pure functions, temporary
    allocation removal,
    constant folding, Numpy \texttt{ufunc} fusion and parallelization, explicit
    thread-level parallelism through OpenMP annotations, false variable
    polymorphism pruning, and automatic vector instruction generation such as
    AVX or SSE.

    In addition to these compilation steps, Pythran provides a C++ runtime 
    library that
    leverages the C++ STL to provide generic containers, and the Numeric 
    % FIXME (phrase trop longuuuuuuuuue) 
    Template Toolbox (NT2) for Numpy support. It takes advantage of modern C++11
    features such as variadic templates, type inference, move semantics and
    perfect forwarding, as well as classical ones such as expression templates.
    % FIXME Expression template is not a "feature" of the language IMO

    % FIXME  (explain that annotations can be added to the code?)
    The input code remains compatible with the Python interpreter, and output
    code is generally as efficient as the annotated Cython equivalent, if not
    more, % FIXME (cut sentence) 
    without the backward compatibility loss of Cython. Numpy expressions
    run faster than when compiled with numexpr or numba, without any change to
    the original code.

@@

The Python language is growing in popularity as a language for scientific
computing, thanks to % FIXME (sounds strange for a publication)
 its concise syntax, a high level standard library and
several scientific packages.

However, the overhead of running a scientific application written in Python
compared to the same algorithm written in a statically compiled language such
as C is high, % FIXME cut sentence
due to numerous dynamic lookup and interpretation cost inherent
in high level languages. Additionally, the Python compiler does not perform
optimization on the byte code, while scientific applications are first-class
candidates for many of them.

Since scientific applications are often told to spend 90\% of their time in
10\% of the code, it is natural to focus on those computation-intensive pieces 
of code. The aim % FIXME (of what) 
may not be to optimize the full Python application, but rather a subset of the application.

Several tools have been developed by an active community to fill the performance
gap encountered when running these computation-intensive piece of code, either
% FIXME after? (perfs issues are not on the compilation (except for...))
through static compilation or Just In Time (JIT) compilation.

One approach, used by Cython \cite{cython2010} is to suppress the interpretation
overhead by translating Python Programs to C programs calling the Python C
API\cite{pythoncapi}. More recently, Nuitka\cite{nuitka}  has taken the same
approach using C++ as a back-end. Going a step further Cython also provides an
hybrid C/Python language that can efficiently be translated into C code. It 
relies both on the Python C API and on plain C.
ShedSkin\cite{shedskin2006} translates implicitly strongly typed Python programs
into C++, without any call to the Python C API, but fails to compile dynamic
codes or codes for which static type inference is impossible.

The alternate approach consists in writing a JIT compiler, embedded into the
interpreter, to dynamically turn the computation intensive parts into native
code. This is what the \texttt{numexpr} module\cite{numexpr} does for Numpy 
expressions, JIT-compiling them from a string representation to native code.
Numba\cite{numba} and Parakeet\cite{parakeet2012} extend this approach to
Numpy-centric applications while PyPy\cite{pypy2009} applies it to the whole
language - including the Numpy module through the in-progress \texttt{numpypy}
branch.

Except for PyPy, none of these compilers apply any of the
static optimization techniques that have been known and successfully applied 
% FIXME And Numba? It generates LLVM, which uses static optimization techniques
for decades to statically compiled language such as C or C++.
Translators to statically compiled languages do take advantage of them
indirectly, but the quality of the generated code may prevent advanced
optimizations, such as vectorization, %FIXME too long
while they are available at higher level,
i.e. at the Python level. Taking into account the specificities of the Python
language can unlock many new transformations. %FIXME sentence without subject
For instance, PyPy automatically
converts the \texttt{range} builtin into \texttt{xrange} by using a dedicated
structure called \texttt{range-list}.

@@

Pythran is a compiler for a subset of the Python language. In this paper, the
name \textbf{Pythran} will be used indifferently to refer to the language or
the associated compiler. The input of the Pythran compiler is a Python module
---not a Python program--- meant to be turned into a native module. Typically,
computation-intensive parts of the program are moved to a module fed to
Pythran. The module is then translated to polymorphic C++ code through the
generation of templated classes in place of Python functions. These
classes are either used from regular C++ % FIXME (not clear what do you mean?)
code or instantiated for given types to generate a native Python module.

It is a strong requirement for Pythran to maintain backward compatibility with
Python, that is, unlike Cython, all code compilable by Pythran is still
executable by a standard-conforming Python interpreter. For this reason, in
addition to language restrictions detailed in the following, Pythran
understands special comments such as:

@@

as an optional module signature. One does not need to list all the module
functions in an \texttt{export} directive, only the functions meant to be used
outside of the module, i.e imported by another module.
Polymorphic functions can be listed several times with
different types. These type annotations are only used for native code generation.

@@

    \item turn Python code into Python Abstract Syntax Tree (AST) by using the \texttt{ast}
   module from the standard library;

@@

Pythran IR is similar to the Python AST, as defined in the \texttt{ast} module, except
that several nodes are forbidden (for instance because Pythran does not support
user-defined classes, or the \texttt{exec} instruction), and some nodes are converted
to others to form a simpler and easier to deal with AST for further analyses and
optimizations. The transformations applied by Pythran on Python AST are the
following:

@@

    \item the closure of nested functions is statically computed to turn nested
        functions into global functions taking the closure elements as
        parameters, and the initial definition is replaced by a \emph{partial
        function} creation, using the \texttt{partial} type from the standard \texttt{functools} module;
        % FIXME (this sentence makes my head hurt :s maybe cut into smaller sentences)

@@

    \item all imports are fully expanded to make functions and global variables access paths explicit,

@@

    \item \texttt{try \dots finally} constructs are turned into nested \texttt{try \dots except} blocks;
        % FIXME (Is the difference between construct and block obvious for intended readers?)

    \item identifiers whose name may be in conflict with C++ keywords are renamed.

@@

The back-end operates in three steps:

@@

The first step requires to map polymorphic variables and polymorphic functions
from the Python world to C++. Pythran only supports polymorphic variables for
functions, i.e. a variable can hold several function objects during its life
time, but it cannot be assigned to a string if it has already been assigned to
an integer. % FIXME (not sure if clear enough) 
As shown later, it is possible to detect several false variable
polymorphism cases using use-def chains. Function polymorphism is achieved
through template parameters: a template function can be applied to several
types as long as an implicit structural typing is respected, which is very
similar to Python's duck typing, except that it is checked at compile time, as
% FIXME (tooooooooooooooooooooooo long) 
illustrated by the following implementation of a generic dot product in Python:

@@

Although far more verbose % FIXME  (don't think you can use verbose this way)
than the Python version, the C++ version also uses some
form of structural typing: the only assumptions these two versions make are that
\texttt{L0} and \texttt{L1} are iterable, their content can be multiplied and 
that the result of the multiplication is accumulatable.

Finally, Pythran computes the set of functions and types used for the processed
module, and generates the minimal set of headers that need to be included to
provide the definitions needed to compile the generated C++ code. This step is
critical to prevent typical slowdown during C++ compilation.

The second step only consists in the instantiation of the top-level functions of
the module, using user-provided signatures. Template instantiation then triggers
the different correctly typed instantiations for all functions written in the
module. % FIXME (not only the functions called by top-level functs????)
module. Note that the user only needs to provide the type of the functions
exported outside the module. The possible types of all internal functions are
then inferred from the call sites. Note that this complex step leverage the 
C++11 compiler capabilities: Pythran generates only one line using user-provided
signature to trigger the template instanciation by the C++ compiler.

The last step involves a template library, called \texttt{pythonic} that 
contains a polymorphic implementation of many functions of the Python standard
library in the form of C++ template functions. Several optimizations, most
notably using expression template, are delegated to this library. Pythran relies
on the C++11\cite{isocxx11} language and leverage recent features such as
move semantics, type inference through \texttt{decltype(\dots)} and variadic
templates.
As a consequence it requires a C++11 compatible compiler for the native code
generation. Boost.Python\cite{boostpython2007} is used for the automatic
Python-to-C++ boilerplate. Generated C++ code is compatible starting with g++ 
4.7.2 and clang++ 3.2.

It is important to note that all Pythran analyses and optimizations are
type-agnostic, i.e. they do not assume any type for the variables manipulated
by the program. The generated C++ template is still type agnostic, the specialization occurs only by generating the template instanciation.
In other words, the Pythran compiler analyzes polymorphic functions and turns
Python modules into C++ meta-programs.

@@

A code analysis is a function that takes a part of the IR (or the whole
module's IR) as an input and returns aggregated high-level information. For
instance, a simple Pythran analysis called \texttt{Identifiers} gathers the set
of all identifiers used throughout the program. This information is later used
when creating new identifiers so that no conflict occurs with already existing ones.

@@

As we do not specialise the analysis for different types and the truth value of
\texttt{a} is unknown at compilation time, the alias analysis yields % FIXME (sounds strange) 
the approximated result that \texttt{c} may point to a fresh location, 
\texttt{a} or \texttt{b}.

Without this kind of information, even a simple instruction like
\texttt{sum(a)} would yield very few information as there is no guarantee that
the \texttt{sum} identifiers point to the \texttt{sum} built-in, as it could have been 
previously stated that \texttt{sum = len}.

When turning Python AST into Pythran IR, nested functions are turned into global
functions taking their expanded closure as parameter. This closure is computed 
using the information provided by the \texttt{Globals} analysis which 
statically computes the state of the dictionary of globals, and 
\texttt{ImportedIds} that computes the set of % FIXME (too long)
identifiers used by an instruction but not declared in this instruction. For
instance in the following snippet:

@@

A rather high-level analysis is the \texttt{PureFunctions} analysis, that computes the
set of functions declared in the module that are pure, i.e. whose return value
only depends on the value of their arguments. This analysis depends on two
other analyses, namely \texttt{GlobalEffects} that computes for each function whether
this function modifies the global state (including I/O, random generators, etc.)
and \texttt{ArgumentEffects} that computes for each argument of each function 
whether or not this argument may be updated in the function body. These three
analyses work inter-procedurally, as illustrated in the following example:

@@

The \texttt{fibo} function is pure as it has no global effect or argument effect and
only calls itself. As consequence the \texttt{bar} function is also pure since the
\texttt{map} intrinsic is pure when its first argument is pure. However the \texttt{foo}
function is not pure since it calls the \texttt{sample} function from the \texttt{random}
module, which has a global effect (on the underlying random number generator
internal state).

Several analyses depend on the \texttt{PureFunctions} analysis.
\texttt{ParallelMaps} uses aliasing information to check if an identifier
points to the \texttt{map} intrinsic, and checks if the first argument is a
pure function using \texttt{PureFunctions}. In that case the \texttt{map} is
added to the set of parallel maps, because it can be executed in any order.
This is the case for the first \texttt{map} in the following snippet, but not
for the second because \texttt{print b} involves an \textbf{I/O}.

@@

\texttt{ConstantExpressions} uses function purity to decide whether or not a given expression
is constant, i.e. its value only depends on literals. For instance the
expression \texttt{fibo(12)} is a constant expression because \texttt{fibo} is pure and its
argument is a literal. 
% FIXME (pourrait etre interressant de repondre a : et t'en fait quoi calcul a la compilation? calcul une fois si appele plusieurs fois ?... ou ref a la transformation qui fait ca)

\texttt{UseDefChains} is a classical analysis from the static compilation world. For
each variable defined in a function, it computes the chain of \textbf{use} and \textbf{def}.
The result can be used to drive various code transformations, for instance to
remove dead code, such as a \textbf{def} followed by aeither nothing or a 
\textbf{def} is useless. 
It is used in Pythran to avoid false polymorphism. An intuitive way to represent
use-def chains is illustrated on the next code snippet:

@@

The fact that all paths finish by a \textbf{def} indicates that the last
assignment can be removed (but not necessarily its right hand part that could
have a side-effect) % FIXME example?
The fact that the last \textbf{use} is always preceded by
a \textbf{def} that only points to it indicates that these \textbf{def}s can
safely be renamed from \texttt{a} to \texttt{a}' (thus performing scalar
renaming).
% FIXME The scalar renaming part is unclear

All the above analyses are used by Pythran developers to build code
transformations that improve the execution time of the generated code.

@@

However, there are some information available at the Python level that cannot
be recovered at the C++ level. For instance, Pythran uses functor with an
internal state and a goto dispatch table to represent generators. This approach
is not very efficient, especially for trivial cases.
Such trivial cases appear when a generator expression is converted, in the
front-end, to a looping generator. To avoid this extra cost, Pythran turns
generator expressions into call to \texttt{imap} and \texttt{ifilter} from the
\texttt{itertools} module whenever possible, removing the unnecessary goto
dispatching table. This kind of transformation cannot be made by the C++
compiler. For instance, the one-liner \texttt{len(set(vec[i]+i for i in cols))}
extracted from the \texttt{nqueens} benchmarks from the Unladen Swallow project
is rewritten as \texttt{len(set(itertools.imap(lambda i: vec[i]+i,cols)))}.
This new form is less efficient in pure Python (it implies one extra function
call per iteration), but can be compiled into C++ more efficiently than a
generic generator.

@@

The information regarding constant expressions is used to perform a classical
transformation called \texttt{ConstantFolding}, which consists in the compile-time
evaluation of constant expressions. The validity is guaranteed by the
\texttt{ConstantExpressions} analysis, and the evaluation relies on Python ability to
compile an AST into byte code and run it, benefiting from the fact that Pythran
IR is a subset of Python AST. A typical illustration is the initialization of a
cache at compile-time:

@@

Sometimes, developers use the same variable in a function to represent value 
with different types, which leads to false polymorphism, as in:

@@

Using the proper library, the C++ language provides an abstraction level close
to what Python proposes. Pythran provides a wrapper library, \texttt{pythonic},
that leverage on the C++ Standard Template Library (STL), the GNU Multiple
Precision Arithmetic Library (GMP) and the Numerical Template Toolbox
(NT2)\footnote{\url{https://github.com/MetaScale/nt2}} to emulate Python
standard library. The STL is used to provide a typed version of the standard
containers (\texttt{list}, \texttt{set}, \texttt{dict} and \texttt{str}), as
well as reference-based memory management through \texttt{shared\_ptr}. Generic
algorithms such as \texttt{accumulate} are used when possible. GMP is the
natural choice to represent Python's \texttt{long} in C++. NT2 provides a generic
vector library called \texttt{boost.simd}\cite{esterie2012boost} that enables
the use of vector instruction units of modern processors in a generic way. It is used
to efficiently compile Numpy expressions.

@@

    \item Expression templates\cite{expression_templates} are used to avoid
        multiple iterations and the creation of intermediate arrays. Because
        they aggregate all \texttt{ufunc} into a single expression at compile
        time, they also increase the computation intensity of the loop body,
        and hence increase the impact of the two following optimizations.

@@

When benchmarked with the \texttt{timeit} module on an hyper-threaded quad-core 
i7, the pure Python execution yields: % FIXME sandy-brige? nehalem?...

@@

Many scientific applications can benefit from the parallel execution of their
kernels. As modern computers generally feature several processors and several
cores per processor, it is critical for the scientific application developers to
be able to take advantage of them.

@@

The OpenMP standard\cite{openmp3.1} is a widely used solution for Fortran, C
and C++ to describe loop-based and task-based parallelism. It consists of a few
directives attached to the code, directives that describe parallel loops and parallel code
sections in a shared memory model.

@@

% FIXME (Ca semble tomber a plat, pas de resultats, pas de conclusion...)

@@

All benchmarks presented in this section are run on an hyper-threaded quad-core
i7, % FIXME sandy-brige, nehalem, what?
using examples shipped along with Pythran sources, available at
https://github.com/serge-sans-paille/pythran in the \texttt{pythran/test/cases}
directory. The Pythran version used is the \texttt{HEAD} of the \texttt{scipy2013} branch,
ShedSkin 0.9.2, PyPy 2.0 compiled with the \texttt{-jit} flag, CPython 2.7.3, Cython
0.19.1 and Numexpr 2.0.1. All timings are made using the \texttt{timeit} module,
taking the best of all runs. All C++ codes are compiled with g++ 4.7.3, using
the tool default compiler option, generally \texttt{-O2} plus a few optimizing flags
depending on the target.

Cython is not considered in most benchmarks, because to get an efficient
binary, one needs to rewrite the original code, while all the considered tools
are running the very same Python code that remains compatible with CPython. The
experiment was performed only to obtain a comparison with Numexpr.

@@

It comes at no surprise that all tools get more than decent results on this
benchmark. PyPy generates a code almost as fast % FIXME (dunno if it's efficient) 
as ShedSkin. Altough both
generate C++, Pythran outperforms ShedSkin thanks to a higher level generated
code. For instance all arrays are represented in ShedSkin by pointers to arrays
that likely disturbs the g++ optimizer, while Pythran uses a vector class wrapping
shared pointers.

@@

It seems that compilers have difficulties to take advantage of high level
constructs such as generator expressions, as the overall speedup is not
breathtaking. Pythran benefits from the conversion to \texttt{itertools.imap} here,
while ShedSkin and PyPy rely on more costly constructs. A deeper look at the
Pythran profiling trace shows that more than half of the execution time is
spent allocating and deallocating a \texttt{set} used in the internal loop. There is a
memory allocation invariant that could be taken advantage of there, but none of
the compiler does so.

@@

The speed ups are not amazing % FIXME (not formal enough, same for previous usage) 
for a numerical application. There are two
reasons for these poor speedups. First, the \texttt{hyantes} benchmark makes heavy
usage of trigonometric functions, and there is not much to gain there. Second, and
most important, the benchmark produces a big 2D array stored as a list of list,
hence the application suffers from the heavy overhead of converting them from C++
to Python. Running the same benchmark using Numpy arrays as core containers
confirms this assumption, as illustrated by Table~\ref{tbl:np-hyantes}. This
table also demonstrates the benefits of manual parallelization using OpenMP.

@@

\begin{table}
    \centering
    \begin{tabular}{|l|c|c|c|c|}
    \hline
     Tool    &  CPython    &  Cython  &  Numexpr    & Pythran   \\
    \hline
     Timing  &  192.2ms    &  36.0ms  &    41.2ms   &  17.1ms   \\
    \hline
     Speedup &  x1         &  x5.33   &  x4.67      &  x11.23   \\
    \hline
\end{tabular}
% FIXME (You changed the order of the compilers, even though pythran is still first why?)
\caption{Benchmarking result on the arc distance kernel.}
\label{tbl:arc-distance}
\end{table}

@@

Although Pythran focuses on a subset of Python and its standard library, many
optimizations opportunities are still present. Using a Domain Specific
Language(DSL) approach, one could use rewriting rules to optimize several
Python idioms. For instance, \texttt{len(set(x))} could lead to an optimized
\texttt{count\_uniq} that would iterate only once on the input sequence.
